---
title: "Asiignemtn_ Groupo_ StatisticalML"
author: "Eduardo Camacho"
date: "September 10, 2024"
output: html_document
---


```{r}
#file.choose()
```

```{r, message=FALSE}
library(dplyr)
library(ggplot2)
library(caret)
```


####  Predictor Variables (X)

The predictor variables used in this analysis represent a wide range of patient characteristics and clinical measurements, including:

**Vital Signs:** These include minimum, maximum, and mean values of heart rate, systolic blood pressure (SysBP), diastolic blood pressure (DiasBP), respiratory rate, temperature, and oxygen saturation (SpO2). These are critical indicators of a patient's physiological condition during their ICU stay.

**Demographic Information:** Variables such as GENDER, DOB (date of birth), and ETHNICITY provide insight into the patient’s background.

**Admission Information:** Variables like ADMITTIME, ADMISSION_TYPE, and INSURANCE indicate the circumstances and context of the patient’s hospital admission.

**Diagnosis Information:** The DIAGNOSIS variable captures the primary diagnosis at the time of ICU admission, while ICD9_diagnosis provides the corresponding ICD-9 code.

##### Independent Variables: mimic_train_X.csv

**subject_id**: Unique identifier for each patient.
**hadm_id**: Hospital admission ID, unique for each hospital admission.
**icustay_id**: Unique identifier for each ICU stay.
**HeartRate_Min**: The minimum heart rate recorded during the ICU stay.
**HeartRate_Max**: The maximum heart rate recorded during the ICU stay.
**HeartRate_Mean**: The average heart rate during the ICU stay.
**SysBP_Min**: Minimum systolic blood pressure recorded during the ICU stay.
**SysBP_Max**: Maximum systolic blood pressure recorded during the ICU stay.
**SysBP_Mean**: The average systolic blood pressure during the ICU stay.
**DiasBP_Min**: Minimum diastolic blood pressure recorded during the ICU stay.
**DiasBP_Max**: Maximum diastolic blood pressure recorded during the ICU stay.
**DiasBP_Mean**: The average diastolic blood pressure during the ICU stay.
**MeanBP_Min**: Minimum mean arterial blood pressure recorded during the ICU stay.
**MeanBP_Max**: Maximum mean arterial blood pressure recorded during the ICU stay.
**MeanBP_Mean**: The average mean arterial blood pressure during the ICU stay.
**RespRate_Min**: Minimum respiratory rate recorded during the ICU stay.
**RespRate_Max**: Maximum respiratory rate recorded during the ICU stay.
**RespRate_Mean**: The average respiratory rate during the ICU stay.
**TempC_Min**: Minimum body temperature (in Celsius) recorded during the ICU stay.
**TempC_Max**: Maximum body temperature (in Celsius) recorded during the ICU stay.
**TempC_Mean**: The average body temperature (in Celsius) during the ICU stay.
**SpO2_Min**: Minimum oxygen saturation (SpO2) level recorded during the ICU stay.
**SpO2_Max**: Maximum oxygen saturation (SpO2) level recorded during the ICU stay.
**SpO2_Mean**: The average oxygen saturation (SpO2) level during the ICU stay.
**Glucose_Min**: Minimum blood glucose level recorded during the ICU stay.
**Glucose_Max**: Maximum blood glucose level recorded during the ICU stay.
**Glucose_Mean**: The average blood glucose level during the ICU stay.

**GENDER**: Gender of the patient.
**DOB**: Date of birth of the patient.
**ADMITTIME**: Timestamp of the patient's hospital admission.
**Diff**: Time difference between hospital admission and another event.
**ADMISSION_TYPE**: Type of admission (e.g., Emergency, Elective).
**INSURANCE**: Patient's insurance type.
**RELIGION**: Patient’s religion.
**MARITAL_STATUS**: Marital status of the patient.
**ETHNICITY**: Patient’s ethnicity.
**DIAGNOSIS**: Initial diagnosis at the time of ICU admission.
**ICD9_diagnosis**: The main diagnosis code in the ICD-9 system.
**FIRST_CAREUNIT**: The first ICU care unit the patient was admitted to (e.g., MICU, SICU, TSICU).


#### Target Variables (Y)

Two possible target variables (dependent variables) are of interest in this analysis:

**HOSPITAL_EXPIRE_FLAG:** A binary variable that indicates whether the patient passed away during their hospital stay (1 = expired, 0 = survived). This will be used for classification tasks.

**LOS (Length of Stay):** A continuous variable representing the total number of days a patient stayed in the hospital. This will be used for regression tasks aimed at predicting the duration of the patient’s stay.


##### Dependent Variables



```{r}
df_x = read.csv("C:\\Users\\Eduardo\\Downloads\\mimic_train_x.csv")
df_y = read.csv("C:\\Users\\Eduardo\\Downloads\\mimic_train_y.csv")
```

```{r}
df <- merge(df_x, df_y, by = "icustay_id")
head(df)
```

```{r}
df$DOB <- as.Date(df$DOB)
df$ADMITTIME <- as.POSIXct(df$ADMITTIME, format="%Y-%m-%d %H:%M:%S")

df$AGE <- as.numeric(difftime(df$ADMITTIME, df$DOB, units = "days")) / 365.25

df <- df %>% select(-c(DOB, ADMITTIME))
```

```{r}
summary(df)
```

```{r}
colSums(is.na(df))
```

```{r}
categorical_cols <- c("GENDER", "ADMISSION_TYPE", "INSURANCE", "RELIGION", 
                      "MARITAL_STATUS", "ETHNICITY", "DIAGNOSIS", 
                      "ICD9_diagnosis", "FIRST_CAREUNIT")

df <- df %>% mutate(across(all_of(categorical_cols), ~ as.numeric(as.factor(.))))
```

```{r}
#summary(df)
```

```{r}
df <- subset(df, LOS >= 0 & LOS <= 70)
summary(df$LOS)
```

```{r}
#summary(df)

ggplot(df, aes(x = LOS)) + 
  geom_histogram(binwidth = .5, fill = "blue", color = "black") + 
  theme_minimal() +
  labs(title = "Distribution of Length of Stay", x = "Length of Stay (days)", y = "Count")

ggplot(df, aes(x = factor(HOSPITAL_EXPIRE_FLAG))) + 
  geom_bar(fill = "blue") + 
  theme_minimal() +
  labs(title = "Distribution of Hospital Expire Flag", x = "Hospital Expire Flag", y = "Count")

```


```{r}
plot(df$LOS, df$HeartRate_Mean,
     main = "Scatter Plot: LOS vs HeartRate_Mean",
     xlab = "Length of Stay (LOS) in Days",
     ylab = "Heart Rate Mean",
     col = "blue", pch = 19)

plot(df$LOS, df$SysBP_Mean,
     main = "Scatter Plot: LOS vs SysBP_Mean",
     xlab = "Length of Stay (LOS) in Days",
     ylab = "Systolic Blood Pressure Mean",
     col = "red", pch = 19)

plot(df$LOS, df$RespRate_Mean,
     main = "Scatter Plot: LOS vs RespRate_Mean",
     xlab = "Length of Stay (LOS) in Days",
     ylab = "Respiratory Rate Mean",
     col = "purple", pch = 19)

plot(df$LOS, df$TempC_Mean,
     main = "Scatter Plot: LOS vs TempC_Mean",
     xlab = "Length of Stay (LOS) in Days",
     ylab = "Temperature Mean (C)",
     col = "green", pch = 19)
```

```{r}
boxplot(df$HeartRate_Mean ~ df$HOSPITAL_EXPIRE_FLAG,
        main = "Heart Rate Mean by Hospital Expire Flag",
        xlab = "Hospital Expire Flag (0 = Survived, 1 = Death)",
        ylab = "Heart Rate Mean",
        col = c("lightblue", "lightcoral"))

boxplot(df$SysBP_Mean ~ df$HOSPITAL_EXPIRE_FLAG,
        main = "Systolic Blood Pressure Mean by Hospital Expire Flag",
        xlab = "Hospital Expire Flag (0 = Survived, 1 = Expired)",
        ylab = "Systolic Blood Pressure Mean",
        col = c("lightblue", "lightcoral"))

boxplot(df$TempC_Mean ~ df$HOSPITAL_EXPIRE_FLAG,
        main = "Temperature Mean by Hospital Expire Flag",
        xlab = "Hospital Expire Flag (0 = Survived, 1 = Expired)",
        ylab = "Temperature Mean (C)",
        col = c("lightblue", "lightcoral"))

boxplot(df$RespRate_Mean ~ df$HOSPITAL_EXPIRE_FLAG,
        main = "Respiratory Rate Mean by Hospital Expire Flag",
        xlab = "Hospital Expire Flag (0 = Survived, 1 = Expired)",
        ylab = "Respiratory Rate Mean",
        col = c("lightblue", "lightcoral"))

```


Data Partition

```{r}
set.seed(123)  
df_train <- createDataPartition(df$HOSPITAL_EXPIRE_FLAG, p = 0.8, list = FALSE)
```



